---
title: "Decoding Retail Transactions"
subtitle: "MSDM CEP Proposal Presentation"
author: "Jake Evans"
institute: "IBM 6540, Cal Poly Pomona"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: [serif, custom.scss]
    width: 1600
    height: 900
    footer: "Decoding Retail Transactions — CEP Proposal | Jake Evans"
    transition: slide
    transition-speed: default
    incremental: false
    toc: true
    toc-depth: 1
    toc-title: "Agenda"
    slide-number: true
    scrollable: true
    smaller: true
    code-fold: false
    code-overflow: wrap
    number-sections: false
    number-depth: 5
    embed-resources: true
    css: styles.css
bibliography: references.bib
editor: visual
execute: 
  echo: true
  freeze: auto
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(knitr)

# Read the Kaggle Retail Sales Dataset if present; otherwise simulate an
# equivalent sample so the deck renders reproducibly anywhere.
if (file.exists("retail_sales_dataset.csv")) {
  sales <- read_csv("retail_sales_dataset.csv", show_col_types = FALSE) |>
    rename_with(~ str_replace_all(tolower(.x), "[^a-z0-9]+", "_"))
} else {
  set.seed(6540)
  n <- 1000
  sales <- tibble(
    transaction_id   = 1:n,
    date             = sample(seq(as.Date("2023-01-01"),
                                  as.Date("2023-12-31"), by = "day"),
                              n, replace = TRUE),
    customer_id      = paste0("CUST", sprintf("%03d", 1:n)),
    gender           = sample(c("Female", "Male"), n, replace = TRUE),
    age              = sample(18:64, n, replace = TRUE),
    product_category = sample(c("Beauty", "Clothing", "Electronics"),
                              n, replace = TRUE, prob = c(.30, .35, .35)),
    quantity         = sample(1:4, n, replace = TRUE),
    price_per_unit   = sample(c(25, 30, 50, 300, 500), n, replace = TRUE)
  ) |>
    mutate(total_amount = quantity * price_per_unit)
}

# Precomputed values for inline narrative (keeps inline R free of $ and *)
n_txn      <- nrow(sales)
date_min   <- format(min(sales$date), "%B %Y")
date_max   <- format(max(sales$date), "%B %Y")
avg_txn    <- paste0("$", round(mean(sales$total_amount), 2))
pct_female <- round(mean(sales$gender == "Female") * 100, 1)
```

# Introduction {background-color="#ede8d8"}

## The Problem

:::::: columns
:::: {.column width="55%"}
Specialty retailers collect transaction logs every day, but most of that data never informs a decision:

::: incremental
- Promotions are planned on **intuition**, not evidence
- Marketing spend is spread **evenly** across categories
- No clear picture of **which customers drive revenue**
:::
::::

::: {.column width="45%"}
> "Without analytics, a transaction log is just a receipt pile."

[The result: budget leaks into segments that were never going to convert.]{.fragment .highlight-red}
:::
::::::

::: notes
Open with the pain point: data exists but is unused. Pause after the quote, then click for the red fragment as the punchline.
:::

## Analytics Objective

. . .

**One objective, stated plainly:**

. . .

Determine how **product category** and **customer demographics** (age, gender) relate to **transaction value**, so promotional spend can be targeted at the highest-value segments.

. . .

::: aside
Scope note: this proposal covers one year of point-of-sale transactions from a single specialty retailer.
:::

## Why It Matters {transition="zoom" transition-speed="slow"}

- Targeted promotion is consistently cheaper than blanket discounting[^1]
- Category-level insight feeds directly into **inventory** and **retail media** decisions
- The method generalizes: the same pipeline works for any transactional dataset

[^1]: Blanket discounts erode margin across all segments, including customers who would have purchased at full price.

# Literature Review {background-color="#ede8d8"}

## Theoretical Grounding

::: incremental
- **Buyer behavior theory** [@howard1969] frames purchase outcomes as a function of buyer characteristics and stimulus (category/price) variables
- **Market segmentation** [@wedel2000] justifies splitting customers by demographics before modeling response
- **RFM analysis** [@hughes1994] supports transaction value as a managerially meaningful outcome
:::

::: notes
Demographic segmentation is grounded in established theory rather than being an arbitrary choice.
:::

## Is the Data Adequate?

::::: columns
::: {.column width="50%"}
**What the data covers**

- Full year of transactions
- Demographics on every record
- Category, quantity, unit price
:::

::: {.column width="50%"}
**What it doesn't**

- [No loyalty / repeat-purchase IDs across visits]{.fragment}
- [No marketing-exposure data]{.fragment}
- [Single retailer → limited generalizability]{.fragment}
:::
:::::

. . .

**Conclusion:** The data is useful for identifying customer segments and analyzing transaction value, but additional information, such as panel data and promotion records, would be needed to make any causal claims.

# Methods {background-color="#ede8d8"}

## Data & Sampling

The dataset is the Kaggle **Retail Sales Dataset**, a point-of-sale extract with `r n_txn` transactions recorded between `r date_min` and `r date_max`. The average transaction value is **`r avg_txn`**, and `r pct_female`% of transactions were made by female customers.[^2]

[^2]: Sample statistics are computed inline from the data, so they update automatically if the dataset changes.

```{r}
#| label: peek
#| code-fold: true
#| code-summary: "Show the import check"
glimpse(sales)
```

## Data Wrangling

```{r}
#| label: wrangle
#| code-line-numbers: "1|2-3|4-8"
#| output-location: fragment
sales_clean <- sales |>
  mutate(age_group = case_when(age < 30 ~ "18–29",
                               age < 45 ~ "30–44",
                               TRUE     ~ "45+")) |>
  group_by(product_category, age_group) |>
  summarize(across(c(quantity, total_amount), mean, .names = "avg_{.col}"),
            n = n(), .groups = "drop")
head(sales_clean)
```

::: notes
Walk through the highlighted lines: start with the pipe, explain how `case_when()` creates the age groups, and then show how `group_by()` and `across()` generate the summary. Click once to reveal the output as a fragment.
:::

## Sample Characteristics

```{r}
#| label: tbl-sample
#| echo: false
sales |>
  group_by(product_category) |>
  summarize(
    Transactions = n(),
    `Avg Age` = round(mean(age), 1),
    `Avg Quantity` = round(mean(quantity), 2),
    `Avg Transaction ($)` = round(mean(total_amount), 2)
  ) |>
  kable(col.names = c("Product Category", "Transactions", "Avg Age",
                      "Avg Quantity", "Avg Transaction ($)"))
```

. . .

Electronics transactions are fewer per customer but dominate on value — the first hint that category, not volume, drives revenue.

## Revenue by Category {.smaller}

```{r}
#| label: fig-revenue
#| code-fold: true
#| code-summary: "Show plotting code"
#| fig-width: 12
#| fig-height: 5.5
sales |>
  mutate(month = floor_date(date, "month")) |>
  group_by(month, product_category) |>
  summarize(revenue = sum(total_amount), .groups = "drop") |>
  ggplot(aes(month, revenue, color = product_category)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::dollar) +
  scale_color_manual(values = c("#C4820E", "#1E4D2B", "#6b7a8f")) +
  labs(x = NULL, y = "Monthly revenue",
       color = "Category",
       title = "Monthly revenue by product category") +
  theme_minimal(base_size = 18)
```

## Measures & Analytics Methods

:::::: columns
::: {.column width="50%"}
**Measures**

- *Outcome:* transaction value (`total_amount`)
- *Predictors:* product category, age group, gender
- *Controls:* quantity, month (seasonality)
:::

:::: {.column width="50%"}
**Methods**

::: incremental
1.  Descriptive segmentation (tables + visualization)
2.  Multiple regression on transaction value
3.  Logistic regression on *high-value transaction* (top quartile) membership
:::
::::
::::::

# Wrap-Up {background-color="#ede8d8"}

## Expected Contribution

- A segment-level map of **where transaction value concentrates**
- A reproducible Quarto pipeline the retailer can rerun each quarter
- Evidence-based targeting recommendations for promotional spend

::: notes
Close by tying back to the objective: one question, answered with data the retailer already owns.
:::

## References

::: {#refs}
:::
